Fast and adaptive cointegration based model for forecasting high frequency financial time series

### 2016-07-06 articles-conferences/iscef2016/src/background.tex
Final manuscript

\section{\uppercase{Background}}
\label{sec:background}

\subsection{Integration and Cointegration}\label{sec:coint}\  
Following Johansen \cite{johansen1995} we shall say that a stochastic process
$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
$\varepsilon_t\sim iid(0,\sigma^2)$ holds.

A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
(discrete) differentiation an stationary process is obtained~\cite{banerjee1993};
more precisely, whenever
$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
all $t$.

@@

Vector error correction model (VECM) describe the joint behaviour
of a set of variables and can be derived from the simple Vector Autoregressive model (VAR) \cite{sims1980} model.
The VAR($p$) model is a framework describing the behaviour of a
set of $l$ endogenous and stationary variables as a linear combination of their last $p$
values, where $l,p\in\mathbb{N}$. 
In our case, each one of these $l$ variables is a scalar time series
$y_{\lambda,t}$, $\lambda=1,\dots,l$, and we represent them all together
at time $t$ by the the vector time series:
\begin{equation}
\label{eq:variables}
\mathbf{y}_t = 
\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
\end{equation}
\noindent
Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.

The VAR($p$) model describes the behaviour of a dependent variable in terms of
its own lagged values and the lags of the others variables in the system. The
model with $p$ lags is formulated as the system of $N$:
\begin{align}
\label{eq:var}
\mathbf{y}_t 
= \boldsymbol{\Phi}_1 \mathbf{y}_{t-1} +
  \boldsymbol{\Phi}_2 \mathbf{y}_{t-2} + \dots +
  \boldsymbol{\Phi}_p\mathbf{y}_{t-p} +
  \mathbf{c} + \boldsymbol{\epsilon}_t \nonumber \\
t=p+1,\dots,N,
\end{align}
\noindent where 
$\boldsymbol{\Phi}_1, \boldsymbol{\Phi}_2,\dots,\boldsymbol{\Phi}_p$
are $l\times l$-matrices of real coefficients,
$\boldsymbol{\epsilon}_{p+1},
 \boldsymbol{\epsilon}_{p+2}, \dots, \boldsymbol{\epsilon}_N$ 
are error terms, $\mathbf{c}$ is a constant vector and $N$ is the total
number of samples.

@@

However, the VAR model cannot be used with non-stationary variables. VECM \cite{engle87} is also a linear model and a special form of a VAR model for I(1) variables that are also
cointegrated~\cite{banerjee1993}. If cointegration exists, variable differences
are stationary and they introduce an error correction term which adjusts
coefficients to bring the variables back to equilibrium.

It is obtained re-writing equation (\ref{eq:var}) in terms of the new
variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
The VECM model, expressed in terms those differences, takes the form:
\begin{equation}\label{eq:vec}
\Delta \mathbf{y}_t 
= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
$\boldsymbol{\Omega}$, expressed in terms of the matrices
$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
\begin{align*}
\boldsymbol{\Phi}_i^* 
&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
\boldsymbol{\Omega}
&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
    - \boldsymbol{\phi}_p \right)\,. 
\end{align*}
The following well known properties of the matrix $\boldsymbol{\Omega}$
\cite{johansen1995} will be useful in the sequel:
\begin{itemize}
\item
If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
\item 
If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
full rank, then the time series are not I(1) but stationary.
\item
If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
and the matrix $\boldsymbol{\Omega}$ can be expressed as
$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
and $\boldsymbol{\beta}$ are
$l\times r$ matrices and
$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
\item
The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
$\boldsymbol{\beta}$ is obtained by Johansen procedure~\cite{johansen1988},
whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
\end{itemize}
It is worth noticing that the factorization of the matrix
$\boldsymbol\Omega$ is not unique, since for any $r \times r$
non-singular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
If cointegration exists, then equation (\ref{eq:vec}) can be written
as follows:
\begin{equation}\label{eq:vecfull}
\Delta\mathbf{y}_t 
= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
which is a VAR model but for time series differences.

>>>>

\section{Background}
\label{sec:background}

\subsection{Integration and Cointegration}\label{sec:coint}\  
Following Johansen \citep{johansen1995} we shall say that a stochastic process
$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
$\varepsilon_t\sim iid(0,\sigma^2)$ holds.

A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
(discrete) differentiation an stationary process is
obtained \citep{banerjee1993}; more precisely, whenever
$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
all $t$.

@@

Vector error correction model (VECM) describe the joint behaviour of a set of
variables and can be derived from the simple Vector Autoregressive model (VAR)
\citep{sims1980} model.  The VAR($p$) model is a framework describing the
behaviour of a set of $l$ endogenous and stationary variables as a linear
combination of their last $p$ values, where $l,p\in\mathbb{N}$.  In our case,
each one of these $l$ variables is a scalar time series $y_{\lambda,t}$,
$\lambda=1,\dots,l$, and we represent them all together at time $t$ by the
vector time series:
\begin{equation}
\label{eq:variables}
\mathbf{y}_t = 
\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
\end{equation}
\noindent
Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.

The VAR($p$) model describes the behaviour of a dependent variable in terms of
its own lagged values and the lags of the others variables in the system. The
model with $p$ lags is formulated as the system:
\begin{align}
\label{eq:var}
\mathbf{y}_t 
= \boldsymbol{\Phi}_1 \mathbf{y}_{t-1} +
  \boldsymbol{\Phi}_2 \mathbf{y}_{t-2} + \dots +
  \boldsymbol{\Phi}_p\mathbf{y}_{t-p} +
  \mathbf{c} + \boldsymbol{\epsilon}_t \nonumber \\
t=p+1,\dots,N,
\end{align}
\noindent where 
$\boldsymbol{\Phi}_1, \boldsymbol{\Phi}_2,\dots,\boldsymbol{\Phi}_p$
are $l\times l$-matrices of real coefficients,
$\boldsymbol{\epsilon}_{p+1},
 \boldsymbol{\epsilon}_{p+2}, \dots, \boldsymbol{\epsilon}_N$ 
are error terms, $\mathbf{c}$ is a constant vector and $N$ is the total
number of samples.

@@

However, the VAR model cannot be used with non-stationary variables. VECM
\citep{engle87} is also a linear model and a special form of a VAR model for I(1)
variables that are also cointegrated \citep{banerjee1993}. If cointegration
exists, variable differences are stationary and they introduce an error
correction term which adjusts coefficients to bring the variables back to
equilibrium.

It is obtained re-writing equation (\ref{eq:var}) in terms of the new
variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
The VECM model, expressed in terms those differences, takes the form:
\begin{equation}\label{eq:vec}
\Delta \mathbf{y}_t 
= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
$\boldsymbol{\Omega}$, expressed in terms of the matrices
$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
\begin{align*}
\boldsymbol{\Phi}_i^* 
&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
\boldsymbol{\Omega}
&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
    - \boldsymbol{\phi}_p \right)\,. 
\end{align*}
The following well known properties of the matrix $\boldsymbol{\Omega}$
\citep{johansen1995} will be useful in the sequel:
\begin{itemize}
\item
If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
\item 
If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
full rank, then the time series are not I(1) but stationary.
\item
If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
and the matrix $\boldsymbol{\Omega}$ can be expressed as
$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
and $\boldsymbol{\beta}$ are
$l\times r$ matrices and
$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
\item
The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
$\boldsymbol{\beta}$ is obtained by Johansen procedure~\citep{johansen1988},
whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
\end{itemize}
It is worth noticing that the factorization of the matrix
$\boldsymbol\Omega$ is not unique, since for any $r \times r$
non-singular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
If cointegration exists, then equation (\ref{eq:vec}) can be written
as follows:
\begin{equation}\label{eq:vecfull}
\Delta\mathbf{y}_t 
= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
which is a VAR model but for time series differences.

### 2016-07-06 articles-conferences/iscef2016/src/intro.tex
Final manuscript

\section{Introduction}
\label{sec:introduction}
In finance, it is common to find variables with long-run equilibrium
relationships. This is termed cointegration and reflects the idea that
some sets of variables cannot wander too far from each other. Cointegration
means that one or more linear combinations of these variables are stationary
even though individually they are not. Some models,
such as the Vector Error Correction (VECM) \cite{engle87}, take advantage of
this property and describe the joint behaviour of several cointegrated
variables. VECM introduces this long-run relationship among a set of cointegrated
time series as an error correction term. These time series must be integrated of order 1, denoted I(1), i.e. they become stationary at their first differences. In finance, I(1) time series are very common and 
to introduce cointegration restrictions in models often improves forecasting \cite{duy1998}. Therefore, VECM has been widely adopted in
financial applications \cite{mukherjee1995,seong2013,maysami2000,arestis2001} such as pair trading \cite{herlemont2003} or models with more than two variables \cite{mukherjee1995,engle2004}.

VECM parameters are obtained using the ordinary least squares
(OLS) method \cite{golub1980}. Since OLS involves many calculations, the parameter estimation is computationally expensive when the number of lagged values and
data increases. Moreover, obtaining cointegration vectors 
is also an expensive routine. This is a main limitation to use VECM with stream data with high frequency.
Chen and Lung's BVECM \cite{chen2003} addresses the advantage of distributed processing over conventional rolling window processing.

Our aims were to study if a parallel version of VECM can be used with high frequency stream data.
Our approach was to determine adaptively the number of observations and lags of VECM which
maximise cointegration relations in the past. 
 This search is computationally
expensive because the Johansen
method \cite{johansen1995} is required to find cointegration vectors, which is of order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error Correction (AVECM), 
is to parallelise this search in order to get new parameters before new data arrives.
Model effectiveness is focused on out-of-sample forecast rather than in-sample
fitting. This criterion allows AVECM prediction capability to be
expressed rather than just explaining data history.
The forecast capability of our method was measured using MSE and the Theil's $U$-statistic \cite{theil1966} widely used in economic forecast. Tests were run using four currency rates: Euro (EUR) to
United States Dollar (USD) (EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss
Franc (CHF) (USDCHF) and USD to Japanese Yen (JPY) (USDJPY) with a 10-seconds frequency.

>>>>

\section{Introduction}
\label{sec:introduction}
In finance, it is common to find variables with long-run equilibrium
relationships. This is termed cointegration and reflects the idea that some sets
of variables cannot wander too far from each other. Cointegration means that one
or more linear combinations of these variables are stationary even though
individually they are not. Some models, such as the Vector Error Correction
(VECM) \citep{engle87}, take advantage of this property and describe the joint
behaviour of several cointegrated variables. VECM introduces this long-run
relationship among a set of cointegrated time series as an error correction
term. These time series must be integrated of order 1, denoted I(1), i.e. they
become stationary at their first differences. In finance, I(1) time series are
very common and to introduce cointegration restrictions in models often improves
forecasting \citep{duy1998}. Therefore, VECM has been widely adopted in financial
applications (\cite{mukherjee1995}, \cite{seong2013}, \cite{maysami2000} and
\cite{arestis2001}). VECM has also been used in pair trading \citep{herlemont2003} or
models with more than two variables (\cite{mukherjee1995} and \cite{engle2004}).

VECM parameters are obtained using the ordinary least squares (OLS) method
\citep{golub1980}. Since OLS involves many calculations, the parameter estimation
is computationally expensive when the number of lagged values and data
increases. Moreover, obtaining cointegration vectors is also an expensive
routine. This is a main limitation to use VECM with stream data with high
frequency.  Chen and Lung's BVECM \citep{chen2003} addresses the advantage of
distributed processing over conventional rolling window processing.

Our aims were to study if a parallel version of VECM can be used with high
frequency stream data.  Our approach was to determine adaptively the number of
observations and lags of VECM which maximise cointegration relations in the
past.  This search is computationally expensive because the Johansen method
\citep{johansen1995} is required to find cointegration vectors, which is of order
$O(n^3)$. Therefore, our proposal, called Adaptively Vector Error Correction
(AVECM), is to parallelise this search in order to get new parameters before new
data arrives.  Model effectiveness is focused on out-of-sample forecast rather
than in-sample fitting. This criterion allows AVECM prediction capability to be
expressed rather than just explaining data history.  The forecast capability of
our method was measured using MSE and the Theil's $U$-statistic \citep{theil1966}
widely used in economic forecast. Tests were run using four currency rates: Euro
(EUR) to United States Dollar (USD) (EURUSD), British Pound (GBP) to USD
(GBPUSD), USD to Swiss Franc (CHF) (USDCHF) and USD to Japanese Yen (JPY)
(USDJPY) with a 10-seconds frequency.

### 2016-07-06 articles-conferences/iscef2016/src/methodology.tex
back to normal
Final manuscript

Cointegration vectors can be found applying the Johansen method which uses a
sample of the last historical data. However, VECM assumes cointegration vectors do not change in time.
In fact, the long-run relationships between the time series might change due to
several economic factors that can lead to structural breaks in the cointegration
relationship \cite{gregoryETal1996}. 
In order to show the number of cointegration vectors which depends on 
the amount $L$ of historical data and the number of lags $p$ in the VECM, we used a grid search. We defined a grid of possible values for $L$ and $p$. $L$ goes throughout $[100,5000]$ with a step size of 300 and $p$  throughout $[1,5]$ with step size of 1. The first experiment consisted in determining the number of cointegration vectors for all combinations of $L$ and $p$.
We used four forex rates: EURUSD, GBPUSD, USDCHF and USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of August 2014,
when the New York and London financial markets opened.

Figure~\ref{fig:hists} shows the distribution of the number of cointegration vectors 
given by the Johansen method for different values of $L$ and $p=1$ . This procedure was carried out by a sliding window of historical data moving 1000 times.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.9\textwidth]{img/histCointVectors-offset20520-p-1-freq-10s}
  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
  possible values for $L$ are shown (100, 2000, 3000 and 4000).}
  \label{fig:hists}
\end{figure}

@@

The corresponding PC values for $L = [100, 2000, 3000, 4000]$ and $p=1$ (figure  \ref{fig:hists}) are: $PC = [21.3, 98.7, 79.8,  65.2]$. For the same values of $L$ and $p=3$ (figure  \ref{fig:histsp3}) the percentage of cointegration is $PC = [18.4, 92.5, 68,  52.5]$. From these results we observed that the number of cointegration vectors changed with different values for $L$ and $p$ and the percentage of cointegration depends mainly on $L$.

The goal of our next experiment was to find a relation between this ratio
$PC$ and the performance measure MSE (see equation 
\ref{eq:MSE}). $L$ was defined between $[100,4000]$ with step size 300 and $p$ takes values between $[1,5]$ with step size 1.

\begin{figure}[ht!]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.8\textwidth]{img/MSE-offset20520-p-1-freq-10s}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. Best $L$ found was 1600. Below MSE versus $L$ shows a rapidly decreasing behaviour, founding minimum at PC = 84.3\% }
  \label{fig:cointvsmse}
\end{figure}

Figure~\ref{fig:cointvsmse} shows the relation between MSE and $PC$ and $L$.
We found that better cointegration percentage leads to better
performance accuracy in terms of reducing MSE. We also found that MSE rapidly decreased with increasing $L$.

@@

Our proposal was then a modified version of VECM, with parameter $p$ and the amount of historical data used $L$ obtained at every step. Only the search of $L$ and $p$ was done in a distributed environment, since this is the most expensive routine. Our proposal is called Adaptive Vector Error Correction model (AVECM).
AVECM is detailed in the algorithm ~\ref{alg:AVECM} which summarises our proposal.

@@

\subsection{Model comparison} \label{sec:random}
We compared our proposal, in terms of performance, with the naive forecast of the random walk model and ARIMA. It is still difficult to outperform the random walk model for standard econometric forecasting models \cite{lo2011} despite its simplicity. The random walk model is defined as:

@@

On the other hand, ARIMA is widely used to forecast returns in finance \cite{tsay2005}. A process can be modelled as an ARIMA$(p,d,q)$ model if $\mathbf{x}_t=\Delta^d \mathbf{y}_t $, i.e after differencing $d$ times the time series $\mathbf{y}_t$,  we get an ARMA$(p,q)$. Since we are modelling returns, we used $d=1$.

@@

Forecast performance was evaluated using two different methods which are frequently used in finance:
\begin{description}
\item
{\bf MSE},  Mean Square Error measures the distance between forecasts
and the true values and large deviations from the true value have a
large impact due to squaring forecast error.
\begin{equation}\label{eq:MSE}
\text{MSE} = 
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
\end{equation}
\item {\bf $U$-statistic}, the Theil's $U$-statistic \cite{theil1966} is a unit free measure obtained as the ratio between the root MSE (RMSE) of a model and the RMSE of the naive random walk model. A $U$-statistic less than 1 implies the performance is better than the naive model.
\end{description}

>>>>

Cointegration vectors can be found applying the Johansen method which uses a
sample of the last historical data. However, VECM assumes cointegration vectors
do not change in time.  In fact, the long-run relationships between the time
series might change due to several economic factors that can lead to structural
breaks in the cointegration relationship \citep{gregoryETal1996}.  In order to
show the number of cointegration vectors which depends on the amount $L$ of
historical data and the number of lags $p$ in the VECM, we used a grid search.
We defined a grid of possible values for $L$ and $p$. $L$ goes throughout
$[2,14]$ hours with a step size of 4 hours and $p$  throughout $[1,5]$ with step
size of 1. The first experiment consisted in determining the number of
cointegration vectors for all combinations of $L$ and $p$.  We used four forex
rates: EURUSD, GBPUSD, USDCHF and USDJPY with 10-second frequency. Data started
at 13:00 GMT of the 13th of August 2014, when the New York and London financial
markets opened.

Figure~\ref{fig:hists} shows the distribution of the number of cointegration
vectors given by the Johansen method for different values of $L=[2,6,10,14]$ and
$p=1$ . This procedure was carried out by a sliding window of historical data
moving 1000 times.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.9\textwidth]{img/Fig1}
  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
  possible values for $L$ are shown (2, 6, 10 and 14) hours (1 hour = 360 data
  points).}
  \label{fig:hists}
\end{figure}

@@

The goal of our next experiment was to find a relation between this ratio $PC$
and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
between $[700,1500]$ data points with step size 100 and $p$ takes values between
$[1,5]$ with step size 1.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.8\textwidth]{img/Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. Best $L$ found was 1060. Below MSE versus $L$ shows a rapidly
  decreasing behaviour, founding minimum at PC = 80\% }
  \label{fig:cointvsmse}
\end{figure}

Figure~\ref{fig:cointvsmse} shows the relation between MSE and $PC$ and $L$.  We
found that better cointegration percentage leads to better performance accuracy
in terms of reducing MSE.

@@

Our proposal was then a modified version of VECM, with parameter $p$ and the
amount of historical data used $L$ obtained at every step. Only the search of
$L$ and $p$ was done in a distributed environment, since this is the most
expensive routine. Our proposal is called Adaptive Vector Error Correction model
(AVECM). AVECM is detailed in the algorithm \ref{alg:AVECM} which summarises
our proposal.

@@

\subsection{Model comparison} \label{sec:random}
We compared our proposal, in terms of performance, with the naive forecast of
the random walk model and ARIMA. It is still difficult to outperform the random
walk model for standard econometric forecasting models \citep{lo2011} despite its
simplicity. The random walk model is defined as:

@@

On the other hand, ARIMA is widely used to forecast returns in finance
\citep{tsay2005}. A process can be modelled as an ARIMA$(p,d,q)$ model if
$\mathbf{x}_t=\Delta^d \mathbf{y}_t $, i.e after differencing $d$ times the time
series $\mathbf{y}_t$,  we get an ARMA$(p,q)$. Since we are modelling returns,
we used $d=1$.

@@

Forecast performance was evaluated using two different methods which are
frequently used in finance:
\begin{description}
\item
{\bf MSE},  Mean Square Error measures the distance between forecasts
and the true values and large deviations from the true value have a
large impact due to squaring forecast error.
\begin{equation}\label{eq:MSE}
\text{MSE} = 
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
\end{equation}
\item {\bf $U$-statistic}, the Theil's $U$-statistic \citep{theil1966} is a unit
free measure obtained as the ratio between the root MSE (RMSE) of a model and
the RMSE of the naive random walk model. A $U$-statistic less than 1 implies the
performance is better than the naive model.
\end{description}
